---
title: "Explore and Mine Data"
author: "Julissa Zelaya"
date: "CS5200 Spring 2024"
output:
  pdf_document: default
  pdf: default
subtitle: Practicum II
---

A pharmaceutical company has provided data regarding their products, sales, 
representatives, and customers. In more specific detail, this data contains the
products being sold by the company, the sales representatives and the
territories they work in, the customer buying products within established
countries, and the sales transactions of products. 

```{r library, message=FALSE, warning=FALSE, echo=FALSE}
## Libraries

# Package names
packages <- c("RMySQL", "DBI", "RSQLite", "ggplot2", "dplyr", "knitr", "pryr")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(!installed_packages)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, function(pkg) {
  library(pkg, character.only = TRUE)
}))
```

```{r db_connection, message=FALSE, warning=FALSE, echo=FALSE}
# Database connection 
db_host <- "sql5.freemysqlhosting.net"
db_user <- "sql5699023"
db_password <- "MiN2PMaGD1"
db_name <- "sql5699023"
db_port <- 3306

# Establish connection
mysql <- dbConnect(
  RMySQL::MySQL(),
  dbname = db_name,
  host = db_host,
  port = db_port,
  user = db_user,
  password = db_password
)
```

## Analytical Query I
Top five sales reps with the most sales broken down by year.
```{sql connection=mysql, message=FALSE, warning=FALSE, echo=FALSE}
SELECT year, repName, total_sales, rank
FROM (
    SELECT 
        a.year, 
        a.repName, 
        a.total_sales,
        COUNT(b.repName) + 1 AS rank
    FROM (
        SELECT year, repName, SUM(totalSales) AS total_sales
        FROM rep_facts
        GROUP BY year, repName
    ) AS a
    LEFT JOIN (
        SELECT year, repName, SUM(totalSales) AS total_sales
        FROM rep_facts
        GROUP BY year, repName
    ) AS b ON a.year = b.year AND a.total_sales < b.total_sales
    GROUP BY a.year, a.repName, a.total_sales
) AS ranked
WHERE rank <= 5
ORDER BY year, rank;

```

## Analytical Query II
Total sold per product per quarter.
```{sql connection=mysql, message=FALSE, warning=FALSE, echo=FALSE}
SELECT productName, CONCAT(year, ' Q', quarter) AS year_quarter, SUM(totalUnitsSold) AS total_units
FROM product_facts
GROUP BY productID, year_quarter
ORDER BY productID, year_quarter
```

## Analytical Query III
Number of units sold per product per region.
```{r units_query, message=FALSE, warning=FALSE, echo=FALSE}
units_per_region <- "
SELECT productName, territory AS region, SUM(totalUnitsSold) AS units_sold
FROM product_facts
GROUP BY productName, region
ORDER BY productName, region
"
# Execute query from units_per_region
query_units <- dbGetQuery(mysql, units_per_region)

query_units <- query_units %>%
  mutate(region = factor(region, levels = unique(region))) # keep region order
```

```{r units_visual, message=FALSE, warning=FALSE, echo=FALSE}
# Plot
ggplot(query_units, aes(x = region, y = units_sold, group = productName, color = productName)) +
  geom_line() +
  geom_point() + # Add points for clarity
  theme_minimal() +
  labs(title = "Units Sold Per Product Per Region",
       x = "Region",
       y = "Units Sold",
       color = "Product") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Adjust for readability
```

## Analytical Query IV
Average sales per sales rep over the years
```{r rep_sales_query, message=FALSE, warning=FALSE, echo=FALSE}
sales_per_rep <- "
SELECT repName, year, AVG(totalSales) AS avg_sales
FROM rep_facts
GROUP BY repName, year
ORDER BY repName, year;
"

# Execute sales_per_rep query
rep_sales <- dbGetQuery(mysql, sales_per_rep)
```

```{r rep_sales_visual, message=FALSE, warning=FALSE, echo=FALSE}
# Plot
ggplot(rep_sales, aes(x = year, y = avg_sales, group = repName, color = repName)) +
  geom_line() +  
  geom_point() + 
  theme_minimal() +
  labs(title = "Average Sales Per Sales Rep Over the Years",
       x = "Year",
       y = "Average Sales",
       color = "Sales Representative") +
  scale_x_continuous(breaks = round(seq(min(rep_sales$year), max(rep_sales$year), by = 1),1)) # Adjust x-axis
```


```{r close_db, message=FALSE, warning=FALSE, echo=FALSE,}
# Close database connection
close <- dbDisconnect(mysql)
```

